---
title: "Lab 02: Data wrangling and visualization"
author: ""
date: "2018-01-25"
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/sta199-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to lab

---

## Agenda

1. **Lab introduction:** Review FiveThirtyEight article on college majors.
2. **Creating team repos:** Accept assignment and join with your teammates.
3. **Starting a new project:** Clone the repo into a new RStudio Cloud project.
4. **Housekeeping & Warm-up:** Make sure your git config and packages are set up.
5. **Tips:** Review some common RMarkdown mistakes.
---

class: center, middle

# Lab introduction

---

## Lab introduction

- Lab instructions posted on the course website under `</>`
    + Let's go find today's lab!
    
- [The Economic Guide To Picking A College Major](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/) by Ben Casselman
    
---

class: center, middle

# Creating team repos

---

## Creating team repos

- Make sure that you have a team name.

- Pick one team member to be the first person to click on the assignment link. This person will enter your team name as they accept the invitation.

- Once their repo is set up, the remaining team members can also accept their assignments and join their team. This gives everyone in a given team the right to push to that single repo.
  
---

class: center, middle

# Starting a new project

---

## Starting a new project

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and hit OK.
---

class: center, middle

# Housekeeping & Warm-up

---

## Housekeeping & Warm-up

- If you didn't do so last week, follow the instructions to configure your git credentials. If you can't remember, just run the following commands and make sure the info is correct:
```
git config --global user.email
git config --global user.name
```
- Install the `tidyverse` package and make sure it's being loaded in your Lab .Rmd.

- Update the name of your project to "Lab 02 - Data wrangling and visualization"

- Pick a team member to walk through the warm-up commit
---

class: center, middle

# Tips

---

## Tips

- Be ready to troubleshoot your document, since it will likely fail to knit on multiple occasions throughout the process. Read the error message carefully and take note of which line is preventing a successful knit.

- Make sure to keep track of your various chunks and to keep text and code in the right place.

- Remember that your .Rmd is not aware of your project's global environment and can only make use of variables, functions, etc. that you have loaded or defined in the document.

- Remind yourself how the pipe operator (`%>%`) works.

- If you're unsure how a function works or what its arguments are, simply type `?` in front of it and hit enter (`?read_csv` for instance). The "Help" tab will open and provide a summary of the function as well as some examples.