---
title: "Lab 09: Simulation based inference"
author: ""
date: "2018-03-29"
output: 
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/sta199-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Agenda

- Next week midterm -- some tips and guidelines
- Lab 09 overview

---

## Midterm 02

- Assigned at noon on Monday, due at noon on Friday
- On Monday: (at a minimum)
    - create repo
    - clone repo to RStudio
    - read the questions
    - make a plan for when you'll work on the exam
- Asking questions on Slack:
    - DM TA + Prof
    - We won't be reading/debugging code
- Exam content:
    - Mostly modelling and inference
    - Some EDA

---

## Lab 09: Simulation based inference

**Data:** Random sample of births from North Carolina

**Goals:** 

- Understand risk factors for infant health
- Use simulation based testing to estimate and test

---

## Setting a seed

- We'll be randomly sampling, so make sure to set a seed.

- All members of a given team should use the same seed to get the same results.

---

## Lab flow

- Part 1: Baby weights

- Part 2: Baby weight vs. smoking

- Part 3: Baby weight vs. mother’s age

Each part requires some data wrangling and visualization along with the inference component

---

## Getting started with lab

- Go to the course GitHub organization and find the lab-09 repo that has your team name on it.

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and hit OK.

- Follow the instructions to configure your git credentials. To double check, run the following 
commands and make sure your info is correct:

```
git config --global user.email
git config --global user.name
```

- The packages you need are already installed
