---
title: "Project tips"
author: ""
date: "2018-04-19"
output: 
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/sta199-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Tips

- Merge conflics will happen, issues will arise, and that’s fine! Commit and push often, and ask questions when stuck.

- Review the grading guidelines on the project assignment and ask questions if any of the expectations are unclear.

- Make sure each team member is contributing, both in terms of quality and quantity of contribution.

- Set aside time to work together and apart (physically).

- When you’re done, review the .md document on GitHub to make sure you’re happy with the final state of your work. Then go get some rest! 
  - If embedding a Shiny app in your app, you should use HTML output.

---

## Code in write up

- In your write up your code should be hidden (`echo = FALSE`) so that your document is neat and easy to read. 

- However your document should include all your code such that if we re-knit your Rmd file we should be able to obtain the results you presented. 

- Exception: If you want to highlight something specific about a piece of code, you’re welcomed to show that portion.

---

## Teamwork

- You are to complete the assignment as a team. 

- All team members are expected to contribute equally to the completion of this assignment and group assessments will be given at its completion - anyone judged to not have sufficient contributed to the final product will have their grade penalized. 

- While different teams members may have different backgrounds and abilities, it is the responsibility of every team member to understand how and why all code and approaches in the assignment work.

---

### RStudio Cloud feedback

- What do you like about being able to access RStudio in the cloud?

- Any pain points (about RStudio Cloud, not the R language)?

---

### Course evaluations

- Please take a few minutes to go on DukeHub and evaluate the course.

- You will receive a separate email about TA evaluations.
